\LARGE \bf Mean Estimation from Adaptive One-bit Measurements

### 2017-10-09 Writeups/onebit_adaptive_Allerton_V3.tex
small changes to final version

In setting (i) the estimator can evaluate the empirical mean of the samples and then communicate it using $n$ bits. This strategy leads to MSE behavior of $\sigma^2/n + O(2^{-n})$. Therefore, the ARE in this setting is $1$. Namely, asymptotically, there is no loss in performance due to the communication constraint under centralized encoding. In this work we show that a similar result does not hold even in setting (ii): the ARE of any adaptive estimation scheme is at least $\pi/2$. Namely, the single-bit per sample constraint incurs a minimal penalty of at least $1.57$ in the number of samples compared to an unconstrained estimator or to the optimal estimator in setting (i). In addition to this negative statement, we provide an estimator that attains this ARE. In other words, we show that the lower bound of $\pi/2$ on the ARE is tight, and that it is attained regardless of the particular realization of $\theta$ or the radius of the parameter space from which it is taken. Clearly, the minimal penalty on the efficiency of $\pi/2$ also holds under setting (iii), although the question whether this efficiency is achievable (or otherwise, what is the minimal ARE) remains open. 
\\

The lower bound of $\pi/2$ on the ARE, i.e., a lower bound of $\sigma^2\pi/(2n) +O(n^{-2})$ on the MSE, is obtained by showing that the Fisher information of any $n$ adaptive messages is not larger than $2n/(\pi \sigma^2)$. From here, the desired bound on the MSE follows from the van Trees version of the information inequality \cite{gill1995applications}. Finally, we show that an estimator that attains asymptotic MSE of $\sigma^2\pi/(2n)$ is obtained as a special case of \cite[Thm. 4]{polyak1992acceleration}. In addition to these two results, we also derive the one-step optimal strategy in which the message sent at step $i$th is designed to minimize the MSE given this message and the previous $i-1$ messages. Furthermore, we demonstrates numerically that the MSE under this strategy converge to $\pi \sigma^2/(2n)$. %A proof for this converges remains elusive. \\

\subsection*{Background and Related Works}

As the variance $\sigma^2$ goes to zero, the task of finding $\theta$ using one-bit queries in the adaptive setting (ii) is easily solved by a bisection style method over the parameter space. Therefore, the general case of non-zero variance is a reminiscent of the noisy binary search problem with possibly infinite number of unreliable tests \cite{cicalese2002least, Karp:2007:NBS:1283383.1283478}. However, since we assume a continuous parameter space, a more closely related problem is that of one-bit analog-to-digital conversion of a noisy signal. For example, the sigma-delta modulator (SDM) analog-to-digital conversion \cite{1092194} uses one-bit threshold detector combined with a feedback loop to update an accumulated error state, and therefore falls under setting (ii). A SDM with a constant input $\theta$ corrupted by a Gaussian noise was studied in \cite{53738}, where it was shown that the output of the modulator converges to the true constant input almost surely. In other words, the SDM provides a consistent estimator for setting (ii). The rate of this convergence, however, was not analyzed and cannot be derived from the results of \cite{53738}. Our results imply that the rate of convergence of a SDM to a constant input is at most $\sigma^2\pi/2$ over the number of feedback iterations. \\

We note that even though its ARE is $1$, the centralized encoding setting (i) already poses a non-trivial challenge for the design and analysis of an optimal encoding and estimation scheme. Indeed, the standard technique to encode an unknown random quantity using $n$ bits is equivalent to the design of a scalar quantizer \cite{gray1998quantization}. However, the optimal design of this quantizer depends on the distribution of its input, which is the goal of our estimation problem and hence its exact value is unknown. As a result, a non-trivial exploration exploitation tradeoff arises in this case. In our setting, the only missing parameter is the mean, which, under setting (i), is known to the encoder with uncertainty interval proportional to $\sigma/\sqrt{n}$. Therefore, while it is clear that uncertainty due to quantization decreases exponentially in the number of bits $n$ leading to ARE $1$, an exact expression for the MSE in this setting is not easy to derive. The situation is even more involved in the adaptive encoding of setting (ii): an encoding and estimation strategy that is optimal for $n-1$ adaptive one-bit messages of a sample of size $n-1$, may not lead to a globally optimal strategy upon the recipient of the $n$th sample. Conversely, any one-step optimal strategy, in the sense that it finds the best one-bit message as a function of the current sample and the previous $2^{n-1}$ messages, is not guaranteed to be globally optimal. Therefore, the main contribution of this paper is a lower bound on the MSE risk of any globally optimal strategy. \\

>>>>

In setting (i), the estimator can evaluate the empirical mean of the samples and then communicate it using $n$ bits. This strategy leads to MSE behavior of $\sigma^2/n + O(2^{-n})$. Therefore, the ARE in this setting is $1$. Namely, asymptotically, there is no loss in performance due to the communication constraint under centralized encoding. In this work we show that a similar result does not hold even in setting (ii): the ARE of any adaptive estimation scheme is at least $\pi/2$. Namely, the single-bit per sample constraint incurs a minimal penalty of at least $1.57$ in the number of samples compared to an unconstrained estimator or to the optimal estimator in setting (i). In addition to this negative statement, we provide an estimator that attains this ARE. In other words, we show that the lower bound of $\pi/2$ on the ARE is tight, and that it is attained regardless of the particular realization of $\theta$ or the radius of the parameter space from which it is taken. Clearly, the minimal penalty on the efficiency of $\pi/2$ also holds under setting (iii), although the question whether this efficiency is achievable (or otherwise, what is the minimal ARE) remains open. 
\\

The lower bound of $\pi/2$ on the ARE, i.e., a lower bound of $\sigma^2\pi/(2n) +O(n^{-2})$ on the MSE, is obtained by showing that the Fisher information of any $n$ adaptive messages is not greater than $2n/(\pi \sigma^2)$. From here, the desired bound on the MSE follows from the van Trees version of the information inequality \cite{gill1995applications}. Finally, we show that an estimator that attains asymptotic MSE of $\sigma^2\pi/(2n)$ is obtained as a special case of \cite[Thm. 4]{polyak1992acceleration}. In addition to these two results, we also derive the one-step optimal strategy in which the message sent at step $i$th is designed to minimize the MSE given this message and the previous $i-1$ messages. Furthermore, we demonstrates numerically that the MSE under this strategy converge to $\pi \sigma^2/(2n)$.\\

We note that even though its ARE is $1$, the centralized encoding setting (i) already poses a non-trivial challenge for the design and analysis of an optimal encoding and estimation scheme. Indeed, the standard technique to encode an unknown random quantity using $n$ bits is equivalent to the design of a scalar quantizer \cite{gray1998quantization}. However, the optimal design of this quantizer depends on the distribution of its input, which is the goal of our estimation problem and hence its exact value is unknown. As a result, a non-trivial exploration exploitation tradeoff arises in this case. Note that the only missing parameter in our setting is the mean, which, under setting (i), is known to the encoder with uncertainty interval proportional to $\sigma/\sqrt{n}$. Therefore, while it is clear that uncertainty due to quantization decreases exponentially in the number of bits $n$ leading to ARE $1$, an exact expression for the MSE in this setting seems to be difficult to derive. The situation is even more involved in the adaptive encoding of setting (ii): an encoding and estimation strategy that is optimal for $n-1$ adaptive one-bit messages of a sample of size $n-1$, may not lead to a globally optimal strategy upon the recipient of the $n$th sample. Conversely, any one-step optimal strategy, in the sense that it finds the best one-bit message as a function of the current sample and the previous $2^{n-1}$ messages, is not guaranteed to be globally optimal. Our results imply that the ARE of any globally optimal strategy is $\pi/2$.

\subsection*{Related Works}

As the variance $\sigma^2$ goes to zero, the task of finding $\theta$ using one-bit queries in the adaptive setting (ii) is easily solved by a bisection style method over the parameter space. Therefore, the general case of non-zero variance is a reminiscent of the noisy binary search problem with possibly infinite number of unreliable tests \cite{cicalese2002least, Karp:2007:NBS:1283383.1283478}. However, since we assume a continuous parameter space, a more closely related problem is that of one-bit analog-to-digital conversion of a noisy signal. For example, the sigma-delta modulator (SDM) analog-to-digital conversion \cite{1092194} uses one-bit threshold detector combined with a feedback loop to update an accumulated error state, and therefore falls under setting (ii). A SDM with a constant input $\theta$ corrupted by a Gaussian noise was studied in \cite{53738}, where it was shown that the output of the modulator converges to the true constant input almost surely. In other words, the SDM provides a consistent estimator for setting (ii). The rate of this convergence, however, was not analyzed and cannot be derived from the results of \cite{53738}. Our results imply that the rate of convergence of the MSE in SDM to a constant input is at most $\sigma^2\pi/2$ over the number of feedback iterations. \\
